---
title: "Tidyverse"
author: "Divij Sinha"
format: html
---

# Tidyverse Examples

## Install and Load the libraries

```{r setup}
# install.packages("tidyverse") ## when/where to do?
library(tidyverse)
setwd("~/Documents/projects/course-website")

knitr::opts_knit$set(root.dir = "~/Documents/projects/course-website")
```

## Read data in

`csv` data from local

```{r}
df <- read_csv("sample_data.csv")
df <- read_csv("week 2/code/sample_data.csv")
df <- readr::read_csv("week 2/code/sample_data.csv")
df <- tidyverse::read_csv("week 2/code/sample_data.csv")
```

excel data

```{r}
df <- readxl::read_excel("week 2/code/sample_data.xlsx")
```

Data from the textbook in `csv`

```{r}
df <- read_csv("https://raw.githubusercontent.com/DataScienceForPublicPolicy/data-sets/refs/heads/master/data/doe_usgulf.csv")
df
```

Data from the textbook in `.Rda`

```{r}
# URL : https://github.com/DataScienceForPublicPolicy/data-sets/blob/master/data/chicago_crime.Rda

# load("https://github.com/DataScienceForPublicPolicy/data-sets/raw/refs/heads/master/data/chicago_crime.Rda")

load("week 2/code/chicago_crime.Rda")
```


## Simple manipulations 

```{r}
budget <- c(
    "Captain’s Log, Stardate 1511.8. I have $10.20 for a big galactic mac.",
    "The ensign has $1,20 in her pocket.",
    "The ExO has $0.25 left after paying for overpriced warp core fuel.",
    "Chief medical officer is the high roller with $53,13."
)


new <- str_replace_all(budget, ",", "\\.")
funds <- str_extract_all(new, "\\$\\d{1,2}\\.\\d{2}", simplify = TRUE)
funds <- str_replace_all(funds, "\\$", "")
funds <- str_trim(funds)
print(funds)
money <- as.numeric(funds)
print(paste0("Total galactic big mac funds = $", sum(money)))

```

Extras: explore `lubridate` for working with dates

Extras: think about Personally Identifiying Information

Extras: Record Linkage

## Tidying data

```{r}
billboard %>%
    pivot_longer(
        cols = starts_with("wk"),
        names_to = "week",
        values_to = "rank"
    ) %>%
    mutate(
        week = parse_number(week)
    )
```

## Cleaning data

```{r}
# install.packages("nycflights13")
library(nycflights13)
flights
```

**What is the average delay time for each airline?**

```{r}
# Calculate average delay time for each airline
avg_delay <- flights %>%
    group_by(carrier) %>%
    summarise(
        avg_arr_delay = mean(arr_delay, na.rm = TRUE),
        avg_dep_delay = mean(dep_delay, na.rm = TRUE)
    )

print(avg_delay)
```

**Which day of the week has the highest number of flights?**

```{r}
# Add a day of the week column
flights <- flights %>%
    mutate(day_of_week = wday(time_hour, label = TRUE))

# Calculate the number of flights for each day of the week
flights_per_day <- flights %>%
    group_by(day_of_week) %>%
    summarise(num_flights = n())

print(flights_per_day)
```

**What is the total number of flights for each destination?**

```{r}
# Calculate the total number of flights for each destination
flights_per_dest <- flights %>%
    group_by(dest) %>%
    summarise(num_flights = n())

print(flights_per_dest)
```

**Which flights have the longest departure delays?**

```{r}
# Find the flights with the longest departure delays
longest_dep_delays <- flights %>%
    filter(!is.na(dep_delay)) %>%
    arrange(desc(dep_delay)) %>%
    head(10)

print(longest_dep_delays)
```

**What is the proportion of flights that are delayed by more than 30 minutes?**

```{r}
# Calculate the proportion of flights delayed by more than 30 minutes
delay_proportion <- flights %>%
    mutate(delayed_30 = ifelse(arr_delay > 30, 1, 0)) %>%
    summarise(proportion_delayed = mean(delayed_30, na.rm = TRUE))

print(delay_proportion)
```


## EDA

```{r}
glimpse(flights)
```

```{r}
summary(flights)
```

```{r}
# Summary statistics
summary_stats <- flights %>%
    summarise(
        total_flights = n(),
        avg_arr_delay = mean(arr_delay, na.rm = TRUE),
        avg_dep_delay = mean(dep_delay, na.rm = TRUE),
        avg_air_time = mean(air_time, na.rm = TRUE),
        avg_distance = mean(distance, na.rm = TRUE)
    )

print(summary_stats)

summary_stats <- flights %>%
    summarise(
        total_flights = n(),
        avg_arr_delay = mean(arr_delay),
        avg_dep_delay = mean(dep_delay),
        avg_air_time = mean(air_time),
        avg_distance = mean(distance)
    )

print(summary_stats)
```

```{r}
missing_values <- flights %>%
    summarise_all(~ sum(is.na(.)))

print(missing_values)

missing_frac <- flights %>%
    summarise_all(~ sum(is.na(.)) / n())

print(missing_frac)
```

## ggplot

```{r}

# 1. Distribution of Arrival Delays
ggplot(flights, aes(x = arr_delay)) +
    geom_histogram(binwidth = 10, fill = "blue", color = "black") +
    labs(title = "Distribution of Arrival Delays", x = "Arrival Delay (minutes)", y = "Frequency")

# 2. Number of Flights per Carrier
flights_per_carrier <- flights %>%
    group_by(carrier) %>%
    summarise(num_flights = n())

ggplot(flights_per_carrier, aes(x = carrier, y = num_flights)) +
    geom_bar(stat = "identity", fill = "green", color = "black") +
    labs(title = "Number of Flights per Carrier", x = "Carrier", y = "Number of Flights")

# 3. Average Delay per Month
avg_delay_per_month <- flights %>%
    group_by(month) %>%
    summarise(
        avg_arr_delay = mean(arr_delay, na.rm = TRUE),
        avg_dep_delay = mean(dep_delay, na.rm = TRUE)
    )

ggplot(avg_delay_per_month, aes(x = factor(month), y = avg_arr_delay)) +
    geom_line(group = 1, color = "red") +
    geom_point(color = "red") +
    labs(title = "Average Arrival Delay per Month", x = "Month", y = "Average Arrival Delay (minutes)")

# 4. Relationship between Air Time and Distance
ggplot(flights, aes(x = distance, y = air_time)) +
    geom_point(alpha = 0.3) +
    labs(title = "Relationship between Air Time and Distance", x = "Distance (miles)", y = "Air Time (minutes)")

# 5. Departure Delays by Carrier
ggplot(flights, aes(x = carrier, y = dep_delay)) +
    geom_boxplot(fill = "orange", color = "black") +
    labs(title = "Departure Delays by Carrier", x = "Carrier", y = "Departure Delay (minutes)")

```


```{r}
flights %>%
    slice_sample(prop = 0.01) %>%
    ggplot(., aes(x = distance, y = air_time)) +
    geom_point(alpha = 0.3) +
    labs(title = "Relationship between Air Time and Distance", x = "Distance (miles)", y = "Air Time (minutes)")


flights %>%
    slice_sample(n = 1000) %>%
    ggplot(., aes(x = distance, y = air_time)) +
    geom_point(alpha = 0.3) +
    labs(title = "Relationship between Air Time and Distance", x = "Distance (miles)", y = "Air Time (minutes)")


flights %>%
    ggplot(., aes(x = distance, y = air_time)) +
    geom_smooth() +
    labs(title = "Relationship between Air Time and Distance", x = "Distance (miles)", y = "Air Time (minutes)")

flights %>%
    ggplot(., aes(x = distance, y = air_time)) +
    geom_smooth(method = "lm") +
    labs(title = "Relationship between Air Time and Distance", x = "Distance (miles)", y = "Air Time (minutes)")


flights %>%
    ggplot(., aes(x = distance, y = air_time)) +
    geom_point(alpha = 0.3) +
    geom_smooth(method = "lm") +
    labs(title = "Relationship between Air Time and Distance", x = "Distance (miles)", y = "Air Time (minutes)")


flights %>%
    slice_sample(n = 10) %>%
    ggplot(., aes(x = distance, y = air_time)) +
    geom_point(alpha = 0.3) +
    geom_smooth(method = "lm") +
    labs(title = "Relationship between Air Time and Distance", x = "Distance (miles)", y = "Air Time (minutes)")


flights %>%
    slice_sample(n = 100) %>%
    ggplot(., aes(x = distance, y = air_time)) +
    geom_point(alpha = 0.3) +
    geom_smooth(method = "lm") +
    labs(title = "Relationship between Air Time and Distance", x = "Distance (miles)", y = "Air Time (minutes)")
```


## Maps!

```{r}
# install.packages("sf")
library(sf)

load("week 2/code/chicago_crime.Rda")

crimes_sf <- crimes %>%
    filter(!is.na(latitude) & !is.na(longitude)) %>%
    st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

ggplot(crimes_sf) +
    geom_sf() +
    labs(title = "Chicago Crime Data", x = "Longitude", y = "Latitude")
```

```{r}
crimes_sf <- crimes %>%
    filter(!is.na(latitude) & !is.na(longitude)) %>%
    filter(latitude > 38) %>%
    st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

ggplot(crimes_sf) +
    geom_sf() +
    labs(title = "Chicago Crime Data", x = "Longitude", y = "Latitude")
```

```{r}
# install.packages("tigris")
library(tigris)
options(tigris_use_cache = TRUE)
chicago_tracts <- tracts(state = "IL", county = "Cook", cb = TRUE)
chicago_tracts
```

FIPS code - Federal Information Processing Series

```{r}
# Join the crime data with the census tracts
crime_tracts <- st_join(crimes_sf, chicago_tracts, join = st_within)
```

CRS - Coordinate Reference System

```{r}
# Join the crime data with the census tracts
crime_tracts <- st_join(crimes_sf, chicago_tracts %>% st_transform(4326), join = st_within)

crime_tracts %>%
    group_by(GEOID) %>%
    summarize(n = n())

crime_tracts %>%
    st_drop_geometry() %>%
    group_by(GEOID) %>%
    summarize(n = n()) %>%
    right_join(chicago_tracts, by = join_by(GEOID)) %>%
    ggplot(data = .) +
    geom_sf(aes(fill = n))


crime_tracts %>%
    st_drop_geometry() %>%
    group_by(GEOID) %>%
    summarize(n = n()) %>%
    left_join(x = chicago_tracts, y = ., by = join_by(GEOID)) %>%
    ggplot(data = .) +
    geom_sf(aes(fill = n))
```

```{r}
crime_tracts %>%
    st_drop_geometry() %>%
    group_by(GEOID) %>%
    summarize(n = n()) %>%
  arrange(desc(n))
```


```{r}

```

