---
title: "simple causal inference"
format: html
---

## Trust in the Police in Belgium (European Social Survey, Round 5)


essround
a numeric for the ESS round

edition
a character for the edition number of the fifth round

idno
a numeric id number

cntry
a character vector for the country (i.e. Belgium, or BE)

trstplc
a numeric vector for trust in the police on an 11-point scale. Higher values indicate more trust. 0 = "no trust at all". 10 = "complete trust"

agea
a numeric vector for the respondent's age

female
a numeric vector for whether the respondent is a woman or not.

eduyrs
a numeric vector for years of education.

hincfel
a numeric vector for the respondent's feeling about their household income. 1 = "living comfortably", 2 = "coping on present income", 3 = "difficult on present income", 4 = "very difficult on present income"

plcpvcr
a numeric vector for how successful police are at preventing crimes in a country on an 11-point scale. 0 = "extremely unsuccessful". 10 = "extremely successful."

```{r}
library(stevedata)
library(tidyverse)
library(cobalt)
df <- ESSBE5 %>% select(-c(cntry, essround, edition)) %>% drop_na()
```



## Do women trust the police more/less than men?

Treatment variable - `female`  
Outcome variable - `trstplc`

## Is the assignment truly random? - balance tests

```{r}
df %>% pivot_longer(c(agea, eduyrs, hincfel, plcpvcr),  names_to = "variable", values_to = "value") %>%
ggplot() +
  geom_histogram(aes(x = value), bins = 30) +
  facet_wrap(variable~female, scales = "free", ncol = 2)
```


```{r}
t.test(agea ~ female, data = df)
```

```{r}
bal.tab(female ~ agea + eduyrs + hincfel + plcpvcr, data = df)
```


### Regression

```{r}
lm(trstplc ~ female + agea + eduyrs + hincfel, data = df) %>% 
  summary()
```
